Controlling synchronization in coupled area-preserving maps using stickiness

### 2018-10-05 Syncarea_control/Das_arxiv/paper.tex
paper updated

\title{Mechanism to control synchronization in coupled area-preserving maps}

@@

\pacs{PACS here}

@@

A surprising feature of stickiness has emerged in our earlier 
work~\cite{Mahata2016}, we have looked at the effects of the mixed phase space 
on the synchronization in a system of two standard maps coupled in 
unidirectional drive-response configuration. We have shown that 
synchronization of chaotic trajectories of the drive and response maps 
typically occurs in the neighborhood of regular inlands as a consequence of 
stickiness in the region.  This is the first instance, as far as we are aware, 
where stickiness have been found to influence synchronization in coupled 
chaotic systems. It is important to point out that a possible role of stickiness in 
Hamiltonian systems for synchronization was already predicted by 
Zaslavsky~\cite{Zaslavsky2002b} in beginning of the last decade.  For a chaotic 
orbit, synchronization typically happens via an intermittent behavior in the 
phase difference of the drive and response maps.   The sticky neighborhoods of 
a regular islands {\it temporarily} traps chaotic orbits. In such trapping 
regions in the phase space, parts of a  chaotic trajectory are almost regular 
in time and allow for synchronization to occur. Such traps are characterized 
using the properties of the finite-time Lyapunov exponent \cite{Szezech2005}. 
In the context of our work, we will refer to these traps as synchronization 
traps. We further note that the behavior of the synchronization traps can be 
analyzed in a more quantitative way by analyzing the location and stability 
properties of the periodic orbits in the vicinity of the locations where 
synchronization takes place.

Synchronization in coupled chaotic systems, coupled map lattices and networks 
have been studied in details (for reviews, see 
\cite{Pecora1998,Boccaletti2002, Pecora2015}). 
Coupled Hamiltonian systems are also known to show synchronization such as 
measure synchronization~\cite{Hampton1999,Wang2003,Vincent2005,Gupta2017} and 
identical synchronization~\cite{Mahata2016,Das2017} However, the ability to control synchronization based on the dynamics of the process has not been given much attention, until 
recently~\cite{Grabow2011,Wang2016}. In realistic systems, the speed of 
synchronization may be significant. For example, in
neurosciences,  the speed of the visual and olfactory processing are 
interesting problems \cite{Thorpe1996,Uchida2003}. In this paper, we intend 
to develop a couple of 
mechanisms to control synchronization process to wit,  to increase and to 
decrease synchronization times based on the location of synchronization traps 
in the sticky neighborhood of regular islands in the phase space of the 
area-preserving system of the standard map. We demonstrate the mechanism at a specific value of the nonlinearly parameter (see Sec.~\ref{Master stability function}) where only two small regular islands exist in the phase space. 
\begin{figure}[t]
	\includegraphics[scale=.45]{Standard_Map}
	\caption{\label{fig:Standard_map} \footnotesize The phase space of the standard map at the parameter value K = 6.0. A couple of regular islands (shown in the color red) and a connected chaotic region (shown in the color blue). }
\end{figure}
The rest of the paper is organized in the following way: we explain the 
coupling scheme in sec.~\ref{sec:Pecora_Carroll} and the mechanism based on 
the location of synchronization traps is given in sec.~\ref{Master stability 
function}. The numerical procedures to delay and advance synchronization 
times are demonstrated in Sec.~\ref{sec:delay} and Sec.~\ref{sec:advanced} 
respectively, and the paper ends with conclusions in 
Sec.~\ref{sec:conclusions} including discussion on limitations and 
implications.

@@

\begin{figure}[t]
	\includegraphics[scale=.3]{pecora_caroll_schematic.eps}
	\caption{\label{fig:schematic} \footnotesize Schematic displaying the Pecora-Carroll method of coupling two standard map as explained in the text.}
\end{figure}
\noindent Here the subscript $n$ denotes the discrete time and $K$ is the 
nonlinearity parameter. These equations typically describe the evolution of 
two canonical variables $P$ and $Q$ which correspond to the momentum and 
co-ordinate in the Poincar\'{e} section of a freely moving rotator with 
interleaved periodic kicks. This system represents the behavior of a variety 
of systems such as charged particle confinement in  mirror magnetic traps, 
particle dynamics in accelerator, comet dynamics in solar systems 
etc.~\cite{Chirikov1960,Izraelev1980,Chirikov1989,Chirikov2008}
Two-dimensional  phase space plots  of the standard map for the parameter 
value $K = 6$ using a chaotic and a few regular initial conditions  are shown in Fig.~\ref{fig:Standard_map}.

@@

We plot the variation in the time-5 finite time Lyapunv exponent (FTLE) defined above at $K = 6$. At the point of synchronization, the FTLE values attain a set of smaller values consistently in a small window, indicating the existence of a synchronization trap.  In Fig.~\ref{fig:FTLE} , the three plots show the fluctuations  of FTLE near the point of synchronization for three different sets of initial conditions - $(P^d_0,Q^d_0,Q^r_0) = (0.466,0.098,0.141)$ in (a), $(0.009,0.026,0.129)$ in (b), and $(0.553,0.554,0.396)$ in (c) with 
synchronization times 287090, 234521, and 164782 respectively. The values on 
the $x$-axis indicate time steps at which averages are computed so that 
synchronization times and the temporal neighborhood are effectively captured 
in the plots wherein a temporary drop is clearly visible, indicated by 
ellipses in red. Synchronization of chaotic trajectories typically occurs in 
one of such traps.

@@

\begin{acknowledgments}
I would like to thank Neelima Gupte and Arnd B\"acker for their suggestions to improve the manuscript. 
\end{acknowledgments}

>>>>

\title{Controlling synchronization in coupled area-preserving maps using 
stickiness}

@@

\pacs{05.45.Xt, 05.45.Ac}

@@

A surprising feature of stickiness has emerged in our earlier 
work~\cite{Mahata2016}, we have looked at the effects of the mixed phase space 
on the synchronization in a system of two standard maps coupled in 
unidirectional drive-response configuration. We have shown that 
synchronization of chaotic trajectories of the drive and response maps 
typically occurs in the neighborhood of regular islands as a consequence of 
stickiness in the region.  This is the first instance, as far as we are aware, 
where stickiness have been found to influence synchronization in coupled 
chaotic systems. It is important to point out that a possible role of stickiness in 
Hamiltonian systems for synchronization was already predicted by 
Zaslavsky~\cite{Zaslavsky2002b} in beginning of the last decade.  For a chaotic 
orbit, synchronization typically happens via an intermittent behavior in the 
phase difference of the drive and response maps.   The sticky neighborhoods of 
a regular islands {\it temporarily} traps chaotic orbits. In such trapping 
regions in the phase space, parts of a  chaotic trajectory are almost regular 
in time and allow for synchronization to occur. Such traps are characterized 
using the properties of the finite-time Lyapunov exponent \cite{Szezech2005}. 
In the context of our work, we will refer to these traps as synchronization 
traps. We further note that the behavior of the synchronization traps can be 
analyzed in a more quantitative way by analyzing the location and stability 
properties of the periodic orbits at the locations where 
synchronization takes place.

Synchronization in coupled chaotic systems, coupled map lattices and networks 
have been studied in details (for reviews, see 
\cite{Pecora1998,Boccaletti2002, Pecora2015}). 
Coupled Hamiltonian systems are also known to show synchronization such as 
measure synchronization~\cite{Hampton1999,Wang2003,Vincent2005,Gupta2017} and 
identical synchronization~\cite{Mahata2016,Das2017} However, the ability to control synchronization based on the dynamics of the process has not been given much attention, until 
recently~\cite{Grabow2011,Wang2016}. In realistic systems, the speed of 
synchronization may be significant. For example, in
neurosciences,  the speed of the visual and olfactory processing are 
interesting problems \cite{Thorpe1996,Uchida2003}. In this paper, we intend 
to develop a couple of 
mechanisms to control synchronization process to wit,  to increase and to 
decrease synchronization times based on the location of synchronization traps 
in the sticky neighborhood of regular islands in the phase space of the 
area-preserving system of the standard map. We demonstrate the mechanism at a specific value of the nonlinearly parameter (see Sec.~\ref{Master stability function}) where only two small regular islands exist in the phase space. 
\begin{figure}[t]
	\includegraphics[scale=.45]{Standard_Map.png}
	\caption{\label{fig:Standard_map} \footnotesize The phase space of the standard map at the parameter value K = 6.0. A couple of regular islands (shown in the color red) and a connected chaotic region (shown in the color blue). }
\end{figure}
The rest of the paper is organized in the following way: we explain the 
coupling scheme in sec.~\ref{sec:Pecora_Carroll} and the mechanism based on 
the location of synchronization traps is given in sec.~\ref{Master stability 
function}. The numerical procedures to delay and advance synchronization 
times are demonstrated in Sec.~\ref{sec:delay} and Sec.~\ref{sec:advanced} 
respectively, and the paper ends with conclusions in 
Sec.~\ref{sec:conclusions} including discussion on limitations and 
implications.

@@

\begin{figure}[t]
	\includegraphics[scale=.3]{pecora_caroll_schematic}
	\caption{\label{fig:schematic} \footnotesize Schematic displaying the Pecora-Carroll method of coupling two standard map as explained in the text.}
\end{figure}
\noindent Here the subscript $n$ denotes the discrete time and $K$ is the 
nonlinearity parameter. These equations typically describe the evolution of 
two canonical variables $P$ and $Q$ which correspond to the momentum and 
co-ordinate in the Poincar\'{e} section of a freely moving rotator with 
interleaved periodic kicks. This system represents the behavior of a variety 
of systems such as charged particle confinement in  mirror magnetic traps, 
particle dynamics in accelerator, comet dynamics in solar systems 
etc.~\cite{Chirikov1960,Izraelev1980,Chirikov1989,Chirikov2008}
Two-dimensional  phase space plots  of the standard map for the parameter 
value $K = 6$ using a chaotic and a few regular initial conditions  are shown in Fig.~\ref{fig:Standard_map}.

@@

We plot the variation in the time-5 finite time Lyapunov exponent (FTLE) 
defined above at $K = 6$. At the point of synchronization, the FTLE values 
attain a set of smaller values consistently in a small window, indicating the 
existence of a synchronization trap.  In Fig.~\ref{fig:FTLE} , the three plots 
show the fluctuations  of FTLE near the point of synchronization for three 
different sets of initial conditions - $(P^d_0,Q^d_0,Q^r_0) = 
(0.466,0.098,0.141)$ in (a), $(0.009,0.026,0.129)$ in (b), and 
$(0.553,0.554,0.396)$ in (c) with 
synchronization times 287090, 234521, and 164782 respectively. The values on 
the $x$-axis indicate time steps at which averages are computed so that 
synchronization times and the temporal neighborhood are effectively captured 
in the plots wherein a temporary drop is clearly visible, indicated by 
ellipses in red. Synchronization of chaotic trajectories typically occurs in 
one of such traps.

@@

\begin{acknowledgments}
I would like to thank Neelima Gupte for discussions and Arnd B\"acker for 
suggestions to improve the manuscript.
\end{acknowledgments}
